Dependent Object Types

### 2012-09-09 doc/fool/article.tex
simple changes

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types and abstract type members, as well as its mixture
of nominal and structural typing through the use of refinement
types. It makes no attempt to model inheritance or mixing
composition. The calculus does not model what's currently in Scala: it
is more normative than descriptive.

@@

The calculus does not precisely describe what's currently in Scala. It
is more normative than descriptive. The main point of deviation
concerns the difference between Scala's compound type formation using
{\bf with} and classical type intersection, as it is modeled in the
calculus. Scala, and the previous calculi attempting to model it
conflates the concepts of compound types (which inherit the members of
several parent types) and mixin composition (which build classes from
other classes and traits). At first glance, this offers an economy of
concepts. However, it is problematic because mixin composition and
intersection types have quite different properties. In the case of
several inherited members with the same name, mixin composition has to
pick one which overrides the others. It uses for that the concept of
linearization of a trait hierarchy. Typically, given two independent
traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
$T_1$ would be available via a super-call. All this makes sense from
an implementation standpoint. From a typing standpoint it is more
awkward, because it breaks commutativity and with it several
monotonicity properties.

In the present calculus, we replace Scala's compound types by
classical intersection types, which are commutative. We also
complement this by classical union types. Intersections and unions
form a lattice wrt subtyping. This addresses another problematic
feature of Scala: In Scala's current type system, least upper bounds
and greatest lower bounds do not always exist. Here is an example:
Given two traits
\begin{lstlisting}
  trait A { type T <: A }
  trait B { type T <: B }
\end{lstlisting}
The greatest lower bound of \code@A@ and \code@B@ is approximated by the
infinite sequence
\begin{lstlisting}
  A with B { type T <: A with B { type T <: A with B {
    type T < ...
  }}}
\end{lstlisting}
The limit of this sequence does not exist as a type in Scala.

@@

The DOT calculus is a simple system of dependent
object-types. \figref{dot-one} gives its syntax, reduction rules,
and type assignment rules.

\subsection{Notation} We use standard notational conventions for
sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
element of $X$. 
%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
%introduce a variable which is already bound in the base-set.
We use an
abbreviation for preconditions in typing judgements. Given an
environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.

@@

Two subclasses of terms are values $v$, which consist of just
variables, and paths $p$ which consist of just variables and field
selections.

@@

      \infrule[\textsc{rfn-$\expand$}]
      {\Gamma \ts T \expand_z {\seq D'}}
      {\Gamma \ts T \refine z {\seq D} \expand_z \seq {D'} \tand \seq D}

@@

Rule (\textsc{rfn-$\expand$}) states that a refinement type $T
\expand_z {\seq D}$ expands to the conjunction of the expansion $D'$
of $T$ and the newly added declarations $D$. Rule
(\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
the same declarations as the upper bound $U$ of $L$ in $T$.  Rules
($\tand$-$\expand$) and ($\tor$-$\expand$) states that expansion
distributes through meets and joins.  Rule (\textsc{$\top$-$\expand$})
states that the top type $\top$ expands to the empty set. Rule
(\textsc{$\bot$-$\expand$}) states that the bottom type $\bot$ expands
to the bottom element $\seq{D_\Bot}$ of the lattice of sets of
declarations (recall \figref{dot-decls}).

@@

\subsection{Sugar: Functions}
Like in Scala, we can encode functions as objects with a special
method. We will freely use the following sugar in the remaining of
this paper. Note that the variable $z$ must be fresh.
\begin{align*}
\arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
\fun x S T t &\iff \new z {{\arrow S T}\ldefs{\mi{apply}(x) = t}} z\\
\app f x &\iff \mapp f {\mi{apply}} x\\
\cast T t&\iff \app {\fun x T T x} t
\end{align*}

@@

\subsubsection{Set of Values}
$\relv k \Gamma s T$ defines the set of values that appear to have
type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
T$ simply by being in the store. In addition, it belongs to $\relv k
\Gamma s T$ for $k > 0$, if it defines all type, method and value labels in
the expansion of $T$ appropriately for $j < k$ steps.
\begin{align*}
&\relv k \Gamma s T = \{ y \ | y \in \dom(s) \andl (\\
& (\Gamma \ts T \wfe \andl\\
&\gap \forall j < k,\\
&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
&\gap \Gamma \ts T \expand_y \seq{D},\\
&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
&\gap\gap t_i \in \rele j {\Gamma \envplus{x_i : S}} s U) \andl\\
&\gap (\forall l_i : V \in \seq{D},\\
&\gap\gap v_i \in \relv j \Gamma s V)) \orl\\
& (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2})) \orl\\
& (T = p.L \andl \Gamma \ts p \ni L \typ {S'..U'} \andl \forall j < k, y \in \relv j \Gamma s {U'})\\
&)\}
\end{align*}

>>>>

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types and abstract type members, as well as its mixture
of nominal and structural typing through the use of refinement
types. It makes no attempt to model inheritance or mixin
composition. The calculus does not model what's currently in Scala: it
is more normative than descriptive.

@@

The calculus does not precisely describe what's currently in Scala. It
is more normative than descriptive. The main point of deviation
concerns the difference between Scala's compound type formation using
{\bf with} and classical type intersection, as it is modeled in the
calculus. Scala, and the previous calculi attempting to model it,
conflates the concepts of compound types (which inherit the members of
several parent types) and mixin composition (which build classes from
other classes and traits). At first glance, this offers an economy of
concepts. However, it is problematic because mixin composition and
intersection types have quite different properties. In the case of
several inherited members with the same name, mixin composition has to
pick one which overrides the others. It uses for that the concept of
linearization of a trait hierarchy. Typically, given two independent
traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
$T_1$ would be available via a super-call. All this makes sense from
an implementation standpoint. From a typing standpoint it is more
awkward, because it breaks commutativity and with it several
monotonicity properties.

In the present calculus, we replace Scala's compound types by
classical intersection types, which are commutative. We also
complement this by classical union types. Intersections and unions
form a lattice wrt subtyping. This addresses another problematic
feature of Scala: In Scala's current type system, least upper bounds
and greatest lower bounds do not always exist. Here is an example:
given two traits
\begin{lstlisting}
  trait A { type T <: A }
  trait B { type T <: B }
\end{lstlisting}
The greatest lower bound of \code@A@ and \code@B@ is approximated by the
infinite sequence
\begin{lstlisting}
  A with B { type T <: A with B { type T <: A with B {
    type T < ...
  }}}
\end{lstlisting}
The limit of this sequence does not exist as a type in Scala.

@@

The DOT calculus is a small system of dependent
object-types. \figref{dot-one} gives its syntax, reduction rules,
and type assignment rules.

\subsection{Notation} We use standard notational conventions for
sets. The notation $\seq{X}$ denotes a set of elements $X$. Given
such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
element of $X$. 
%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
%introduce a variable which is already bound in the base-set.
We use an
abbreviation for preconditions in typing judgements. Given an
environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.

@@

Two sub-sorts of terms are values $v$, which consist of just
variables, and paths $p$ which consist of just variables and field
selections.

@@

      \infrule[\textsc{rfn-$\expand$}]
      {\Gamma \ts T \expand_z {\seq {D'}}}
      {\Gamma \ts T \refine z {\seq D} \expand_z \seq {D'} \tand \seq D}

@@

Rule (\textsc{rfn-$\expand$}) states that a refinement type $T
\expand_z {\seq D}$ expands to the conjunction of the expansion $\seq {D'}$
of $T$ and the newly added declarations $\seq D$. Rule
(\textsc{tsel-$\expand$}) states that a type selection $p.L$ carries
the same declarations as the upper bound $U$ of $L$ in $T$.  Rules
($\tand$-$\expand$) and ($\tor$-$\expand$) states that expansion
distributes through meets and joins.  Rule (\textsc{$\top$-$\expand$})
states that the top type $\top$ expands to the empty set. Rule
(\textsc{$\bot$-$\expand$}) states that the bottom type $\bot$ expands
to the bottom element $\seq{D_\Bot}$ of the lattice of sets of
declarations (recall \figref{dot-decls}).

@@

\subsection{Sugar: Functions}
Like in Scala, we can encode functions as objects with a special
method. We will freely use the following sugar in the remaining of
this paper. Note that the variable $z$ must be fresh.
\begin{align*}
\arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
\fun x S T t &\iff \new z {{\arrow S T}\ldefs{\mi{apply}(x) = t}} z\\
\app f x &\iff \mapp f {\mi{apply}} x\\
\cast T t&\iff \app {(\fun x T T x)} t
\end{align*}

@@

\subsubsection{Set of Values}
$\relv k \Gamma s T$ defines the set of values that appear to have
type $T$ when taking at most $k$ steps. $\Gamma$ and $s$ must agree:
$\dom(\Gamma) = \dom(s)$ (ordered) and $\forall (x : T) \in \Gamma, x
\in \relv k \Gamma s T$. A variable $y$ belongs to $\relv 0 \Gamma s
T$ simply by being in the store. In addition, it belongs to $\relv k
\Gamma s T$ for $k > 0$, if it defines all type, method and value labels in
the expansion of $T$ appropriately for $j < k$ steps.
\begin{align*}
&\relv k \Gamma s T = \{ y \ | y \in \dom(s) \andl (\\
& (\Gamma \ts T \wfe \andl\\
&\gap \forall j < k,\\
&\gap y \mapsto T_c \ldefs{\seq{l = v}\;\seq{m(x)=t}} \in s,\\
&\gap \Gamma \ts T \expand_y \seq{D},\\
&\gap (\forall L_i : S \rightarrow U \in \seq{D},\\
&\gap\gap \Gamma \ts y \ni L_i \typ {S'..U'}) \andl\\
&\gap (\forall m_i : S \rightarrow U \in \seq{D},\\
&\gap\gap t_i \in \rele j {\Gamma \envplus{x_i : S}} s U) \andl\\
&\gap (\forall l_i : V \in \seq{D},\\
&\gap\gap v_i \in \relv j \Gamma s V)) \orl\\
& (T = T_1 \tand T_2 \andl y \in \relv k \Gamma s {T_1} \andl y \in \relv k \Gamma s {T_2}) \orl\\
& (T = T_1 \tor T_2 \andl  (y \in \relv k \Gamma s {T_1} \orl y \in \relv k \Gamma s {T_2}))\\
&)\}
\end{align*}
